Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-03-18 POPL/paper.tex
popl/prelim paper stub

\documentclass[sigconf,anonymous,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{stmaryrd}

@@

\begin{abstract}
  We propose a semantic framework for the study of formal
  grammars. First, we provide a syntax-independent notion of formal
  grammar as a function from strings to sets, generalizing the
  familiar notion of a language as a set of strings. The set of such
  grammars naturally form a very rich category whose notion of
  isomorphism corresponds to strong equivalence of grammars, and for
  which many language-theoretic constructs can be defined using
  universal properties.

  Based on these category-theoretic constructions we propose a new
  syntactic formalism for formal grammars: a version of dependent
  linear-non-linear type theory in which the tensor product is not
  commutative. The linear types can be interpreted as grammars, and
  linear terms as parse transformers. The non-dependent fragment of
  this type theory is already enough to express regular/context-free
  expressions as well as finite automata, with the isomorphism between
  regular languages and automata definable as linear terms. This
  generalizes the language theoretic formalisms such as Kleene
  algebras to incorporate ambiguity. By incorporating a dependency of
  linear types on non-linear types, we can express richer grammar
  formalisms such as indexed grammars, pushdown automata and Turing
  machines. This provides a new form of ``logical characterization''
  of grammar classes based on a substructural logic rather than
  ordinary first-order logic. Further, the type theory can not only
  express the grammars as types, but equivalences between grammars and
  parsers as terms of the grammar.

  We give this type theory a semantics in the category of grammars and
  prove a canonicity theorem that shows that every term in a context
  corresponding to a fixed string is equal to a term in a canonical
  form encoding a parse tree of that string.
\end{abstract}

@@

\section{A Syntax-Independent Notion of Syntax}

The theory of formal languages and parsing is one of the oldest and
most thoroughly developed areas of theoretical computer science. A
prominent topic in the 1950s and 60s led to a series of remarkable
developments: Chomsky's hierarchy of grammar formalisms, practical
algorithms for parsing of regular and context-free grammars and
variants, as well as implementations of practical tools for the
generation of efficient parsers.

A central notion is that of a \emph{formal language} $L$ over an
alphabet $\Sigma$ as simply a \emph{subset} of the set of strings $L
\subseteq \Sigma^*$. This definition is especially useful as it gives
a semantics to formal grammars that is completely independent of any
particular syntactic grammar formalism: any new notion of grammar can
be given a semantics in this common framework and it provides a
precise mathematical speicification for implementing a
\emph{recognizer} of a language and comparing the power of different
formalisms by demonstrating what languages can be formalized within
it. This also allows for completely \emph{language theoretic}
formulations of language classes: e.g., the regular \emph{languages}
can be characterized as those that have finitely many derivatives
\cite{brzozowski-or-myhill-nerode-idk}. This notion is remarkable
because it makes no reference to any specific syntactic formalism for
defining languages such as regular expressions or finite automata.

Formal language theory alone is not sufficient as a specification of a
parser: a language $L \subseteq \Sigma^*$ is equivalently defined as
its indicator function $\chi_L : \Sigma^* \to \Prop$ mapping each
string $w$ to the proposition that it is in the language $w \in
L$. This can then serve as the specification for a language
\emph{recognizer}: a program $r$ of type $\Sigma^* \to \Bool$ such
that $r(w) = \texttt{true}$ if and only if $\chi_L(w)$ holds, but
recognition is insufficient for most tasks for which the theory of
parsing was developed: the production of \emph{semantic} structures
from synactic descriptions in linguistics, compilation or
deserialization. The output of a parser is not just a boolean but a
\emph{parse tree}\footnote{such parse trees are usually not
materialized in memory, as the construction of the parse tree is fused
with application of semantic actions, which can be seen as a kind of
fold over the generated tree.}.

Due to this limitation, parsers are typically specified not by a
formal language, but by some \emph{formal grammar}, which specifies
not just \emph{which} strings are in the language but specifies what
are the \emph{parse trees} for the string. However unlike formal
language theory, there is no common universal notion of what
constitutes a formal grammar, but rather many syntactic systems such
as Chomskyan(?)  grammars, Thue systems, Montague grammars, etc which
are all syntactic presentations of the same underlying idea of an
abstract specification for parsing. The most common of these is the
Chomsky hierarchy of \emph{generative} grammars which specify parse
trees using \emph{derivations} of the string from a set of rewrite
rules, the parse tree encoding which rules were used.

Our first contribution is conceptual: we propose a simple,
syntax-independent definition of \emph{formal grammar}:
\begin{definition}
  A \emph{formal grammar} over an alphabet $\Sigma$ is a function
  $\Sigma^* \to \Set$.
\end{definition}
We say that a grammar $G$ associates to every string $w$ the set $Gw$
of \emph{parses} of that string. While we will work informally in this
paper, the collection $\Set$ can be formalized in many different
logical foundations: as a universe in type theory, a proper class in
set theory, etc. We argue that this is already a common intuition
latent in much prior work on grammars, especially when comparing
different formalisms for equivalence such as \cite{??}. This notion is
also completely natural in the context of constructive type theories
such as Agda which until recently did not include a universe of
propositions, and so for example in Elliott (\cite{??}) a formal
language was defined as a function to the universe $\Type$ with little
comment.

Every formal grammar $G$ induces a formal language, which can be seen
most easily by using the characteristic function formulation: we can
``squash'' any set into the proposition that it is inhabited, so
$\chi_G(w) = ||G(w)||$ which defines a subset $L_G = \{ w \in
\Sigma^*| G(w) \mathrm{inhabited}\}$.

Formal languages naturally arrange themselves into a partial order by
using the subset inclusion as the ordering. Formal grammars naturally
arrange themselves into a \emph{category} where a morphism $\alpha : G
\to H$ is given by a \emph{family} of functions
\[ \alpha^w : Gw \to Hw \]
for every string $w \in \Sigma^*$. The intuition is that a morphism of
grammars is a \emph{translation of parses}: a $G$-parse of $w$ gets
transformed to an $H$-parse of $w$. This then induces a notion of
\emph{isomorphism} of grammars: two grammars are equivalent if there
is a bijective translation of parses, precisely capturing the notion
of a \emph{strong equivalence} of grammars.

This category-theoretic framework can be further used to describe and
compare different notions of formal grammar. We can define a
\emph{notion of grammar} to be a category (or perhaps just groupoid)
paired with a functor into $\Set^{\Sigma^*}$. This naturally forms a
2-category the slice 2-category over $\Set^{\Sigma^*}$ and equivalence
in this 2-category defines a sensible notion of \emph{strong
equivalence of grammar formalisms} (TODO: check details).

The category of grammars is quite rich in structure, which can be most
easily observed from the fact that it is equivalently defined as the
category of functors $\Set^{\Sigma^*}$ where $\Sigma^*$ is here viewed
as a discrete category, that is, the objects are strings and only
morphisms are identity morphisms. Such functor categories, often
called presheaf categories, are incredibly rich in structure, which we
will exploit in section \ref{blah} to succinctly define the semantics
of regular and context-free grammars.

\section{Kleene Category}

Kleene algebras are an important tool in the theory of regular
languages. More broadly, they serve as a theoretical substrate to
studying various kinds of formal languages. Formally, they are a tuple
$(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is a set, $+$ and $\cdot$
are binary operations over $A$, $(-)^*$ is a function over $A$, and
$1$ and $0$ are constants. These structures satisfy the axoims depicted
in Figure~\ref{fig:axioms}.

\begin{figure}
  \begin{align*}
    x + (y + z) &= (x + y) + z & x + y &= y + x\\
    x + 0 &= x & x + x &= x\\
    x(yz) &= (xy)z & x1 &= 1x = x\\
    x(y + z) &= xy + xz & (x + y)z &= xz + yz\\
    x0 &= 0x = x & & \\
    1 + aa^* &\leq a^* & 1 + a^*a &\leq a^*\\
     b + ax \leq x &\implies a^*b \leq x &  b + xa \leq x &\implies ba^* \leq x
  \end{align*}
  \caption{Kleene algebra axioms}
  \label{fig:axioms}
\end{figure}

The addition operation can be used to define the partial order
structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
formal languages, this order structure can be used to model language
containment. In this section, we want to categorify the concept of
Kleene algebra and build on top of it in order to define an abstract
theory of parsing.  We start by defining \emph{Kleene categories}.

\begin{definition}
  A Kleene category is a distributive monoidal category $\cat{K}$
  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that $B \otimes (\mu
  X.\, F_{A, 1}) \cong \mu X.\, F_{A, B}(X)$ and the analogue isomorphism
  for $G_{A,B}$ also holds.
\end{definition}

As a sanity check, note that Kleene algebras are indeed examples of
Kleene categories.

\begin{example}
  Every Kleene algebra, seen a posetal category, is a Kleene category.
\end{example}

This example illustrates how restricted Kleene algebras are in terms
of reasoning about languages. By only having at most one morphism
between objects, there is not a lot of information they can convey. In
this case, the only information you get is language containment. We
will see throughout this paper how the extra degrees of freedom
granted by having more morphisms give you more algebraic structure for
reasoning about languages.

For the next example, we see an unexpected connection with the theory
of substructural logics.

\begin{example}
  The opposite category of every Kleene category is a model of a variant of
  conjunctive ordered logic, where the Kleene star plays the role of the ``of
  course'' modality from substructural logics which allows hypotheses to
  be discarded or duplicated.
\end{example}

The proposed axioms are a direct translation of the Kleene algebra
axioms to a categorical setting. Its most unusual aspect is the
axiomatization of the Kleene star as a family of initial algebras
satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
more structure, then these isomorphisms hold ``for free''.

\begin{theorem}
  \label{th:kleeneclosed}
  Let $\cat{K}$ be a Kleene category such that it is also monoidal
  closed.  Then, the initial algebras isomorphisms hold automatically.
\end{theorem}
\begin{proof}
  We prove this by the unicity (up-to isomorphism) of initial
  algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
  (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
  X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
  (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
    1}(X))$.

  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
  we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,1} \to B \lto Y$. We can show existence and
  uniqueness by showing that the diagram on top commutes if, and
  only if, the diagram on the bottom commutes:

% https://q.uiver.app/#q=WzAsOCxbMCwwLCJCICsgQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFsyLDAsIkIgKyBZIFxcb3RpbWVzIEEiXSxbMiwyLCJZIl0sWzAsMiwiQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDMsIjEgKyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDUsIlxcbXUgWC5cXCwgMSArIFggXFxvdGltZXMgQSJdLFsyLDMsIjEgKyAoIEIgXFxsdG8gWSkgXFxvdGltZXMgQSJdLFsyLDUsIkIgXFxsdG8gWSJdLFswLDEsImlkICsgKGlkIFxcb3RpbWVzIGg7IGV2KSBcXG90aW1lcyBpZF9BIl0sWzEsMiwiW2YsZ10iXSxbMywyLCJpZCBcXG90aW1lcyBoOyBldiIsMl0sWzAsMywiW2lkIFxcb3RpbWVzIGgsIGlkIFxcb3RpbWVzIHRsXSIsMl0sWzQsNiwiaWQgKyAoaCBcXG90aW1lcyBYKVxcb3RpbWVzIGlkIl0sWzUsNywiaCIsMl0sWzYsNywiW2YnLCBnJ10iXSxbNCw1LCJbaGQsIHRsXSIsMl1d
\[\begin{tikzcd}
	{B + B \otimes (\mu X.\, 1 + X \otimes A)} && {B + Y \otimes A} \\
	\\
	{B \otimes (\mu X.\, 1 + X \otimes A)} && Y \\
	{1 + (\mu X.\, 1 + X \otimes A)} && {1 + ( B \lto Y) \otimes A} \\
	\\
	{\mu X.\, 1 + X \otimes A} && {B \lto Y}
	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
	\arrow["{[f,g]}", from=1-3, to=3-3]
	\arrow["{id \otimes h; ev}"', from=3-1, to=3-3]
	\arrow["{[id \otimes h, id \otimes tl]}"', from=1-1, to=3-1]
	\arrow["{id + (h \otimes X)\otimes id}", from=4-1, to=4-3]
	\arrow["h"', from=6-1, to=6-3]
	\arrow["{[f', g']}", from=4-3, to=6-3]
	\arrow["{[hd, tl]}"', from=4-1, to=6-1]
\end{tikzcd}\]

  This equivalence follows by using the adjunction structure given
  by the monoidal closed structure of $\cat{K}$. A completely analogous
  argument for $G_{A,B}$ also holds.
\end{proof}

This result feels similar in spirit to the definition of action
algebras, which are algebras where the product also has adjoint
operations which results in the Kleene star being more easily
axiomatized \cite{kozen1994}.

We are now ready to prove that our concept of formal grammars fits
nicely within our categorical framework. We start by presenting a
well-known construction from presheaf categories.

\begin{definition}
  Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
  two functors $\cat{C} \to \Set$. Their Day convolution tensor
  product is defined as the following coend formula:
  \[
  (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z) 
  \]
  Dually, its internal hom is given by the following end formula:
  \[
  (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
  \]
\end{definition}

\begin{lemma}[\cite{day1970}]
  Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
  monoidal closed.
\end{lemma}

%% \begin{theorem}
%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
%%   The functor category $[A, \cat{K}]$.
%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
%% \end{theorem}
\begin{theorem}
  If $\cat{C}$ is a locally small monoidal category, then
  $\Set^{\cat{C}}$ is a Kleene category.
\end{theorem}
\begin{proof}

  By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
  is a presheaf category, it has coproducts. Furthermore, the tensor
  is a left adjoint, i.e. it preserves colimits and, therefore, it is
  a distributive category.

  As for the Kleene star, since presheaf categories admit small colimits,
  the initial algebra of the functors $F_{A,B}$ and $G_{A,B}$ can be
  defined as the filtered colimit of the diagrams:

  From Theorem~\ref{th:kleeneclosed} it follows that these initial
  algebras satisfy the required isomorphisms and this concludes the
  proof.
\end{proof}

\begin{corollary}
  For every alphabet $\Sigma$, the presheaf category $\Set^{\cat{\Sigma^*}}$
  is a Kleene category.
\end{corollary}
\begin{proof}
  Note that string concatenation and the empty string make the
  discrete category $\Sigma^*$ a strict monoidal category.
\end{proof}

Much like in the posetal case, the abstract structure of a Kleene
category is expressive enough to synthetically reason about formal
languages. A significant difference between them is that while Kleene
algebras can reason about language containment, Kleene categories can
reason about parsing as well.

For the rest of the paper we will use the presheaf category
$\Set^{\cat{\Sigma^*}}$ as a concrete model that will serve as our
guide when expanding abstract formalism presented in this section so
that it can handle more classes of languages beyond regular ones.

\paragraph{Beyond Simple Types}

Though Kleene categories are expressive enough to reason about
concepts that are outside of reach of Kleene algebras, their
simply-typed nature makes them not so expressive from a type theory
point of view. This is limiting because type theories are tested
syntactic frameworks for manipulating complicated categories
structures while avoiding some issues common in category theory, such
as coherence issues.

With this in mind, we want to design a dependent type theory that
builds on top of Kleene categories with the goal of internalizing
various reasoning principles for languages and their parsers. This
leads us to the abstract notion of model we are interested in
capturing with our type theory.

\begin{definition}
  A model is a locally Cartesian category with two hierarchies of
  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
  $U_i$ to be closed under dependent products and sum,
  $L_i$ to be closed under the Kleene category connectives, 
  dependent products, left and right closed structures, with
  a type constructor $G : L_i \to U_i$ and a linear dependent sum
  going the other direction.
\end{definition}

\begin{theorem}
  The presheaf category $\Set^{\cat{\Sigma^*}}$ is a model.
\end{theorem}

\section{Related work}

\paragraph{Kleene Algebra}

Since the early works in the theory of formal languages, Kleene
algebras have played an important role in its development. They
generalize the operations known from regular languages by introducing
operations generalizing language composition, language union and the
Kleene star.  More generally, they are defined as inequational theory
where the inequality is meant to capture language containment. This
theory is extremely successful, having found applications in algebraic
path problems, theory of programming languages, compiler optimizations
and more.

A frequently fruitful research direction is exploring varying
extensions of Kleene algebras, Kleene algebra with tests (KAT) being
one of the most notable ones. Our approach is radically different from
most extensions, which usually aim at modifying or adding new
operations to Kleene algebras, but still keeping it as an inequational
theory. By adopting a category-theoretic treatment and allowing the
``order structure'' to encode more information than merely
inequalities, we were able to extend Kleene algebra to reason about
parsing as well.

\paragraph{TODO: more stuff}

%% That paper with Fritz Henglein about regexes as types

%% Vaughan Pratt on the residual operator

%% \section{Universal Constructions in the Category of Grammars}

%% Predicate BI: bicartesian closed, monoidal biclosed, omega-colimits
%% and therefore certain initial algebras

\section{Non-commutative Linear-Non-Linear Type Theory as a Syntax for Grammars}
\label{sec:typetheory}
\subsection{Syntax}
Judgments: $\Delta$ set ctx, $\Delta \vdash A$ set, $\Delta \vdash M :
A$ term, $\Delta \vdash \Gamma$ grammar ctx, $\Delta \vdash G$
grammar, $\Delta\pipe \Gamma \vdash p : G$ parse term.

\begin{figure}
  \label{fig:structjdg}
  \begin{mathpar}
    \inferrule{~}{\ctxwff \cdot}
    \and
    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}

    \\

    \inferrule{~}{\linctxwff \Gamma \cdot}
    \and
    \inferrule{\linctxwff \Gamma \Delta \\ \linctxwffjdg \gamma A}{\linctxwff \Gamma {\Delta, a : A}}

    \\

    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}

    \and

    \inferrule{\Gamma \vdash A : L_i}{\linctxwffjdg \Gamma A}

    \\

    \inferrule{\Gamma \vdash X \equiv Y : U_i}{\ctxwffjdg \Gamma {X\equiv Y}}

    \and

    \inferrule{\Gamma \vdash A \equiv B : L_i}{\linctxwffjdg \Gamma {A \equiv B}}

  \end{mathpar}
  \caption{Structural judgements}
\end{figure}

\begin{figure}
  \label{fig:typewf}
  \begin{mathpar}
    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
 %   
    \and
%
    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
%
    \and
%
    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash 1 : U_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash I : L_i}
 %   
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
    %
    \and
    %
    \inferrule{~}{\Gamma \vdash \top : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
%
    \\
    %
    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
    %
    \and
    %
    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
  \end{mathpar}
  \caption{Type well-formedness}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
%
    \and
%
    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
%
    \and
%
    \inferrule{~}{}
%
    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
%   
    \and
%
    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
%
    \and
%
    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
%
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
%
    \and
%
    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C} 
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
\end{mathpar}
  \caption{Judgemental equality}
\end{figure}

\begin{figure}
  \label{fig:inttyping}
  \begin{mathpar}
  \inferrule{~}{\Gamma, x : X, \Gamma' \vdash x : X}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : Y \quad \ctxwffjdg \Gamma {X \equiv Y}}{\Gamma \vdash e : X}
  %
  \\\
  %
  \inferrule{~}{\Gamma \vdash () : 1}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : X \\ \Gamma \vdash e : \subst Y e x}{\Gamma \vdash (e, e') : \SigTy x X Y}
  %
  \\
%
  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_1\, e : X}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : \SigTy x X Y}{\Gamma \vdash \pi_2\, e : \subst Y {\pi_1\, e} x}
  \and
  \inferrule{\Gamma, x : X \vdash e : Y}{\Gamma \vdash \lamb x e : \PiTy x X Y}
  %
  \and
  %
  \inferrule{\Gamma \vdash e : \PiTy x X Y \\ \Gamma \vdash e' : X}{\Gamma \vdash \app e {e'} : \subst Y {e'} x}
  %
  \\
  %
  \inferrule{\Gamma \vdash e \equiv e' : X}{\Gamma \vdash \mathsf{refl} : e =_X e'}
  \and
  \inferrule{\Gamma \mid \cdot \vdash e : A}{\Gamma \vdash \mathsf G e : \mathsf G A}
  \end{mathpar}
  \caption{Intuitionistic typing}
\end{figure}

\begin{figure}
  \label{fig:linsyntax}
  \begin{mathpar}
    \inferrule{~}{\Gamma \mid a : A \vdash a : A}
    \and
    \inferrule{\Gamma \mid \Delta \vdash e : B \\ \linctxwffjdg \Gamma {A \equiv B}}{\Gamma \mid \Delta \vdash e : A}
    %
    \\
    %
    \inferrule{~}{\Gamma \mid \cdot \vdash () : I}
    \and
    \inferrule{\Gamma \mid \Delta \vdash e : I \\ \Gamma \mid \Delta_1',\Delta_2' \vdash e' : C}{\Gamma \mid \Delta_1',\Delta,\Delta_2' \vdash \letin {()} e {e'} : C}
    %
    \\
    %
    \inferrule{\Gamma \mid \Delta \vdash e : A \\ \Gamma \mid \Delta' \vdash e' : B}{\Gamma \mid \Delta, \Delta' \vdash e \otimes e' : A \otimes B}
    %
    \\
    %
    \inferrule{\Gamma \mid \Delta \vdash e : A \otimes B \\ \Gamma \mid \Delta'_1, a : A, b : B, \Delta'_2 \vdash e'}{\Gamma \mid  \Delta_1', \Delta, \Delta'_2 \vdash \letin {a \otimes b} e {e'}}
    \\
    %
    \inferrule{\Gamma \mid \Delta, a : A \vdash e : B}{\Gamma \mid \Delta \vdash \lamblto a e : A\lto B}
    \and
    \inferrule{\Gamma \mid \Delta' \vdash e' : A \\ \Gamma \mid \Delta \vdash e : A \lto B}{\Gamma \mid \Delta', \Delta \vdash \applto {e'} {e} : B}
    \\
    %
    \inferrule{\Gamma \mid a : A, \Delta \vdash e : B}{\Gamma \mid \Delta \vdash \lambtol a e : B\tol A}
    \and
    \inferrule{\Gamma \mid \Delta \vdash e : B \tol A \\ \Gamma \mid \Delta' \vdash e' : A}{\Gamma \mid \Delta, \Delta' \vdash \apptol e {e'} : B}
    %
    \\
    %
    \inferrule{\Gamma, x : X \mid \Delta  \vdash e : A}{\Gamma \mid \Delta \vdash \dlamb x e : \LinPiTy x X A}
    \and
    \inferrule{\Gamma \mid \Delta \vdash e : \LinPiTy x X A \\ \Gamma \vdash e' : X}{\Gamma \mid \Delta \vdash \app e {e'} : \subst A {e'} x}
    %
    \\
    %
    \inferrule{\Gamma \vdash e : X \quad \Gamma \mid \Delta \vdash e' : \subst A e x}{\Gamma \mid \Delta \vdash (e, e') : \LinSigTy x X A}
    %
    \\
    %
    \inferrule{\Gamma \mid \Delta \vdash e : \LinSigTy x X A \quad \Gamma, x : X \mid \Delta'_1, a : A, \Delta'_2 \vdash e' : C}{\Gamma\mid \Delta'_1, \Delta, \Delta'_2 \vdash \letin {(x, a)} e {e'}: C}
    %
    \\
    %
    \inferrule{~}{\Gamma \mid \Delta \vdash () : \top}
    %
    \\
    %
    \inferrule{\Gamma \mid \Delta \vdash e_1 : A_1 \quad \Gamma \mid \Delta \vdash e_2 : A_2}{\Gamma \mid \Delta \vdash (e_1, e_2) : A_1 \amp A_2}
    \and
    \inferrule{\Gamma \mid \Delta \vdash e : A_1 \amp A_2 }{\Gamma \mid \Delta \vdash \pi_i \, e : A_i}
    %
    \\
    %
    \inferrule{\Gamma \vdash e : \mathsf{G} A}{\Gamma \mid \cdot \vdash \mathsf{G}^{-1}\, e : A}
    %
    \\
    %
    \inferrule{\Gamma; \Delta \vdash e : \subst A {\mu x.\, A} x}{\Gamma; \Delta \vdash \mathsf{cons}\, e : \mu x.\, A}
    \and
    \inferrule{\Gamma;\Delta\vdash e' : \mu x.\,A \and \Gamma; a:\subst A B x \vdash e : B}{\Gamma;\Delta\vdash \mathsf{fold}(a.e)(e') : B}
  \end{mathpar}
  \caption{Linear typing}
\end{figure}

\pedro{Need to figure out the rules for inductive types and write down
  the judgemental equality rules}

\subsection{Semantics in Grammars}

>>>>

% -*- fill-column: 80; -*-
\documentclass[sigconf,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{stmaryrd}

\usepackage{todonotes}
\newcommand{\todoin}[1]{\todo[inline]{TODO:\@ #1}}

@@

\begin{abstract}
  We propose a semantic framework for the study of formal grammars. First, we
  provide a syntax-independent notion of formal grammar as a function from
  strings to sets of parse trees, generalizing the familiar notion of a language
  as a set of strings. This point-of-view naturally endows a grammar with the
  structure of a presheaf, which allows many language-theoretic constructs to be
  defined via universal properties.

  With these categorical constructions in mind, we propose a new syntactic
  formalism for formal grammars: a version of dependent linear-non-linear type
  theory where the tensor product is non-commutative. The linear types in this
  theory are interpreted as grammars --- and linear terms as parse transformers.
  The non-dependent fragment of this type theory is already expressive enough to
  capture regular expressions, context-free expressions, and finite automata.
  Moreover, the isomorphism between regular languages and finite automata is
  definable as a proof term in our formalism. The
  non-dependent fragment of this type theory is expressive enough to capture
  both regular and context-free expressions, as well as finite automata already
  regular expressions, context-free expressions, and finite automata. Moreover,
  the is definable as a proof term in this formalism. Further, through use of
  dependency of linear types on non-linear types, we can express richer grammar
  formalisms such as indexed grammars, pushdown automata, and Turing machines.

  Classically, many results in both language and automata theory are proved
  informally with respect to some ambient foundation --- e.g. Kleene algebra, grammars,
  parsers, semantic actions, etc. The type theory provided in
  this paper serves as an intuitive and general framework to syntactically unify
  these related ideas and internalize their
  theorems formally. We give a new form of \emph{logical characterization} of
  grammar classes based on a substructural logic rather than ordinary
  first-order logic. Further, the type theory can not only
  express the grammars as types, but equivalences between grammars and
  parsers as terms of the grammar.

  We give this type theory a semantics in the category of grammars and
  prove a related canonicity theorem. Moreover, we keep an eye toward the
  implementation of this type theory for correct-by-construction parsers and
  parser combinators.
\end{abstract}

@@

\section{A Syntax-Independent Notion of Syntax}
\todoin{Rework intro from LICS draft with a focus on implementation}

\section{Kleene Categories}
\todoin{Tell a story winding through the use of KAs, action algebras, etc to
  motivate why we'd want categorical semantics for KCats}
\todoin{Incoporate Kleene Category definition}
\todoin{End with the punch that $\Set^{\Sigma^{*}}$ is a KCat}

\section{Automata}
